RVD2: An ultra-sensitive variant detection model for low-depth heterogeneous next-generation sequencing data

### 2015-03-09 doc/bioinformatics2014/revision2/response_to_reviews.txt

Thank you for identifying this important omission. The region of interest for the PAXIP1 gene should have covered the entire 59Kb coding region. We ran RVD2 on the current NCBI start/stop codons (http://www.ncbi.nlm.nih.gov/gene/22976) to give the expanded region of chr7:154,735,400 - 154,794,682. This expanded region yields additional mutations. In particular, using the expanded region, we identified 5 additional germline variants and 2 additional somatic variants. In expanding the region of interest the previously identified germline variants at chr7:154,766,700 and chr7:154,781,769 are not called. A close inspection shows that these loci have very similar point estimates of the allele frequency in the case and control in the shortened region of interest. In the analysis with the expanded region, RVD2 estimates a lower across-locus precision parameter (M0) and thus does not call these loci. The control M0 parameter is 5.7 in the previous analysis and 4.2 in the 59kbp region.

3. Instead of using DNA sequencing of primary tumors genomes, which are highly heterogeneous and present many rare (single nucleotide) variants, the authors used the DNA from a commercially available cell line, which is colonially expanded (genetically homogenous). Interestingly, chromosome 7, where the investigated gene is located, is tetraploid on HCC1187[1], therefore, one would expect to find only allele frequencies 0%, 25%, 50%, 75 and 100% on variants located in this chromosome. Figure 5 seems to be in agreement to this hypothesis, except for variants 49704 and 81769, which has marginally lower estimations of allele frequency.Therefore, the variants detected on HCC1187 are not actually rare variants (MAF=25%) and, as shown on Figure 3 (Table1), may be detected by most of currently available methods. Despite the limitations, the analysis is complete, but it would be more informative to apply RVD2 to publicly available (and validated) DNAseq from primary tumors.

Thank you for pointing out the evidence that HCC1887 is tetraploid for chromosome 7. We have incorporated this evidence into our discussion of the data with the reference your provided. We agree that it would be informaive to apply RVD2 to publicly available and validated DNAseq from primary tumors. We have searched extensively for such a data set and we have experienced three limitations. First, RVD2 requires sequencing data from a control sample. The control sample does not need to be matched, but should be of the same sequence as the case. This control sample allows RVD2 to measure the locus-specific error rate and call significant deviations from that error rate in the case. This requirement limits the pool of possible data sets. Second, we require a reasonable read depth (as shown by the synthetic DNA experiments) to identify rare variants. Many DNAseq data sets have a read depth of 10-15x because the investigators are seeking to identify clear variants. Few data sets have a read depth higher than 15x. Finally, we wanted a data set for which the originators provide a set of validated variants. Again, this requirement limits the number of available data sets. Taken together, these requirements are met by the HCC1187 data set. We expect that with the increased interest in tumor heterogeneity, investigators will be generating more data sets that meet these criteria. It will be interesting to explore those data sets with RVD2.

### 2015-03-09 doc/bioinformatics2014/revision2/document.tex
Responded to reviewer 1 comment 3. Removed list of PAXIP1 mutation calls from narrative.

RVD2 identified twelve variants in the 59kbp PAXIP1 gene from chr7:154,735,400 to chr7:154,794,682. There were 11 germline variants and 10 somatic mutations. %Especially, two heterozygous germline variants had a loss-of-heterozygosity (LOH) in the tumor sample.
Figure~\ref{fig:brca_MAF} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations. 
Positions 
chr7:154,743,899C\textgreater T, 
chr7:154,749,704G\textgreater A, 
chr7:154,753,635T\textgreater C, 
chr7:154,754,371T\textgreater C, 
chr7:154,758,813G\textgreater A, 
chr7:154,780,960C\textgreater T, 
chr7:154,784,369A\textgreater G, 
chr7:154,787,357C\textgreater T, 
chr7:154,787,367C\textgreater T, 
chr7:154,787,612A\textgreater G, and 
chr7:154,788,798A\textgreater G 
were called germline mutations. 
%
Positions 
chr7:154,749,704G\textgreater A, 
chr7:154,753,635T\textgreater C, 
chr7:154,754,371T\textgreater C, 
chr7:154,758,813G\textgreater A, 
chr7:154,760,439A\textgreater C, 
chr7:154,766,732T\textgreater G,  
chr7:154,766,832A\textgreater C,
chr7:154,777,118A\textgreater C,
chr7:154,787,612A\textgreater C,
chr7:154,788,798A\textgreater C, and
were identified as significantly different in tumor and normal sample and called somatic mutations.
% 
Positions chr7:154,754,371 and chr7:154,758,813 appear to be loss-of-heterozygosity events. 
Some of these mutations are also found to be common population SNPs according to dbSNPv138. 
The corresponding identities are shown in the Figure~\ref{fig:brca_MAF}. The read depth distribution for positions called by RVD2 are provided in Supplementary Table 1. 
%According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.

>>>>

RVD2 identified twelve variants in the 59kbp PAXIP1 gene from chr7:154,735,400 to chr7:154,794,682. There were 11 germline variants and 10 somatic mutations. %Especially, two heterozygous germline variants had a loss-of-heterozygosity (LOH) in the tumor sample.
Figure~\ref{fig:brca_MAF} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations. 
%
%Positions 
%chr7:154,743,899C\textgreater T, 
%chr7:154,749,704G\textgreater A, 
%chr7:154,753,635T\textgreater C, 
%chr7:154,754,371T\textgreater C, 
%chr7:154,758,813G\textgreater A, 
%chr7:154,780,960C\textgreater T, 
%chr7:154,784,369A\textgreater G, 
%chr7:154,787,357C\textgreater T, 
%chr7:154,787,367C\textgreater T, 
%chr7:154,787,612A\textgreater G, and 
%chr7:154,788,798A\textgreater G 
%were called germline mutations. 
%%
%Positions 
%chr7:154,749,704G\textgreater A, 
%chr7:154,753,635T\textgreater C, 
%chr7:154,754,371T\textgreater C, 
%chr7:154,758,813G\textgreater A, 
%chr7:154,760,439A\textgreater C, 
%chr7:154,766,732T\textgreater G,  
%chr7:154,766,832A\textgreater C,
%chr7:154,777,118A\textgreater C,
%chr7:154,787,612A\textgreater C,
%chr7:154,788,798A\textgreater C, and
%were identified as significantly different in tumor and normal sample and called somatic mutations.
% 
Interestingly, positions chr7:154,754,371 and chr7:154,758,813 appear to be loss-of-heterozygosity events. 
Some of these mutations are also found to be common population SNPs according to dbSNPv138. 
The corresponding identities are shown in the Figure~\ref{fig:brca_MAF}. The read depth distribution for positions called by RVD2 are provided in Supplementary Table 1. 
%According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.

Interestingly, karyotyping indicates that chromosome 7 in HCC1187 is tetraploid (http://www.path.cam.ac.uk/~pawefish/BreastCellLineDescriptions/HCC1187.html).
